Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-05-31 Manuscript/Imputation_of_Incomplete_Multilevel_Data2.tex
update table 1 and 2 tables in appendix
Please just refer to changes in tables, other changes i did was to solve running problems

\author{
Hanne I. Oberman\\Methodology and Statistics\\
Utrecht University \And Johanna Muñoz\\Julius Center for Health Sciences
and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \And Gerko Vink\\Methodology and Statistics\\
Utrecht University \AND Valentijn M. T. de Jong\\Julius Center for
Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\
Data Analytics and Methods Task Force,\\
European Medicines Agency,\\
Amsterdam, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne I. Oberman, Johanna Muñoz, Thomas P. A. Debray, Gerko
Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}.
}

@@

\Address{
    Hanne I. Oberman\\
    Methodology and Statistics\\
Utrecht University\\
    Padualaan 14\\
3584 CH Utrecht\\
  E-mail: \email{h.i.oberman@uu.nl}\\
  URL: \url{https://hanneoberman.github.io/}\\~\\
          }

@@

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
Sample unit & Units of the population from which measurements are taken
in a sample. \\
Hierarchical levels & Data are grouped into clusters at different
levels. A three-level \\
Fixed effect & Here we assume that the values of an independent variable
are fixed, i.e., \\
& the values observed in the study are representative of all values in
the \\
& in the dependent variable y e.g., blood pressure between treatments A
and B. \\
Random effect & The values of an independent variable are assumed to be
randomly drawn from \\
& admission we might select only certain hospitals that are
representative of \\
& the difference of y between individual hospitals, but rather the
variation of \\
ICC & The variability due to clustering is often measured by means of
the \\
& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
& of variance that can be attributed to the cluster-level, where a
high \\
& ICC would indicate that a lot of variability is due to the cluster \\
& structure. \\
Random effect & Multilevel models typically accommodate for variability
by including \\
& a separate group mean for each cluster. In addition to random \\
& intercepts, multilevel models can also include random coefficients \\
& and heterogeneous residual error variances across clusters {[}see
e.g. \\
& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
stratified intercept as concept..{]} \\
\bottomrule()
\end{longtable}

@@

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/patterns-1}

@@

\caption[Sporadic missingness in multilevel data]{Sporadic missingness in multilevel data}\label{fig:patterns}
\end{figure}
\end{CodeChunk}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
MCAR & Missing Completely At Random, where the probability to be missing
is equal \\
& across all data entries \\
MAR & Missing At Random, where the probability to be missing depends on
observed \\
& information \\
MNAR & Missing Not At Random (MNAR), where the probability to be
missing \\
& depends on unrecorded information, making the missingness
non-ignorable \\
& \citep{rubi76, meng94}. \\
& \\
\bottomrule()
\end{longtable}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)        # for reproducibility
R> library(mice)         # for imputation
R> library(miceadds)     # for additional imputation routines
R> library(ggmice)       # for incomplete/imputed data visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel parameter pooling
R> library(miceheckman)  # for imputation cf. heckman models
R> library(metamisc)     # for case study data
\end{CodeInput}
\end{CodeChunk}

@@

In this section we will go over the different steps involved with
imputing incomplete multilevel data with the R package mice. We consider
the simulated \texttt{popmis} dataset, which included pupils
(\(n = 2000\)) clustered within schools (\(N = 100\)). The following
variables are of primary interest:

@@

@@

First we plot the pattern of missing data within categories of the
relevant variables. Plot the missing data pattern:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(popmis, square = TRUE)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_pat-1}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-corr-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-hist-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-pred-1} \end{center}

@@

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_multilevel-1} \end{center}

@@

We illustrate how to impute incomplete multilevel data by means of a
case study: \texttt{impact} from the \pkg{metamisc} package
\citep[empirical data on traumatic brain injuries, \(n = 11,022\) units
across \(N = 15\) clusters,][]{metamisc}. The \texttt{impact} data set
contains traumatic brain injury data on \(n = 11022\) patients clustered
in \(N = 15\) studies with the following 11 variables:

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification,
\item
  \texttt{hypox} Hypoxia (0=no, 1=yes),
\item
  \texttt{hypots} Hypotension (0=no, 1=yes),
\item
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
\item
  \texttt{edh} Epidural hematoma (0=no, 1=yes),
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/forest-1} \end{center}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pattern-1} \end{center}

@@

This shows that we need to impute \texttt{ct} and \texttt{pupil}.

To develop the best imputation model, we need to investigate the
relations between the observed values of the incomplete variables and
the observed values of other variables, and the relation between the
missingness indicators of the incomplete variables and the observed
values of the other variables. To see whether the missingness depends on
the observed values of other variables, we can test this statistically
or use visual inspection (e.g.~a histogram faceted by the missingness
indicator).

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact_corr-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect  group term  estimate std.error statistic   p.value conf.low conf.high
   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
\end{CodeOutput}
\end{CodeChunk}

As we can see, a higher \texttt{ct} (Marshall Computerized Tomography
classification) is associated with a lower odds of 6-month mortality,
given by the odds ratio exp(0.42), CI \ldots{} to \ldots, when
controlling for\ldots{}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> meth <- make.method(dat) # methods vector
R> pred <- quickpred(dat)   # predictor matrix
R> plot_pred(pred, rotate = TRUE)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-1} \end{center}

\begin{CodeInput}
R> pred[pred == 1] <- 2
R> pred["mort", ] <- 2
R> pred[, "mort"] <- 2
R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
R> pred[, "name"] <- -2
R> diag(pred) <- 0
R> plot_pred(pred, rotate = TRUE)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/impact-2} \end{center}

@@

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10505.026 10570.795 -5243.513 10487.026     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2908  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32327                 -0.42365                  0.03023  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67143                 -1.05776                 -1.51039  
                  pupil                       ct  
                0.49756                  0.42474  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10519.511 10585.280 -5250.755 10501.511     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2961  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.33576                 -0.40872                  0.03039  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66477                 -1.05453                 -1.51860  
                  pupil                       ct  
                0.45928                  0.44419  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

>>>>

\author{
Hanne I. Oberman\\Utrecht University \And Johanna Munoz\\Julius Center
for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care \And Gerko Vink\\Utrecht University
\AND Valentijn M. T. de Jong\\Julius Center for Health Sciences and
Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne I. Oberman, Johanna Munoz, Thomas P. A. Debray, Gerko
Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}. Footnotes in the current
version show work in progress/under construction. The last section is
not part of the manuscript, but purely for reminders. See also all of
the TODOs that need to be worked out. We aim to submit at JSS, so there
is no word count limit (``There is no page limit, nor a limit on the
number of figures or tables''). {[}Just adding some text to get a better
guess of what the actura abstract will look like: Lorem ipsum dolor sit
amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut
labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud
exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
Duis aute irure dolor in reprehenderit in voluptate velit esse cillum
dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non
proident, sunt in culpa qui officia deserunt mollit anim id est
laborum.{]}
}

@@

\Address{
    Hanne I. Oberman\\
    Utrecht University\\
    Padualaan 14\\
3584 CH Utrecht\\
  E-mail: \email{h.i.oberman@uu.nl}\\
  URL: \url{https://hanneoberman.github.io/}\\~\\
        Thomas P. A. Debray\\
    Julius Center for Health Sciences and Primary Care\\
    University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\

@@

Many datasets include individuals from multiple settings, geographic
regions, or even different studies. In the simplest case, individuals
(e.g., students) are nested within so-called clusters (e.g., school
classes). More complex clustered structures may occur when there are
multiple hierarchical levels (e.g., patients within hospitals within
regions or countries), or when the clustering is non-nested (e.g.,
electronic health record data from diverse settings and populations
within large databases). In general, individuals from the same cluster
tend to be more similar than individuals from other clusters. In
statistical terms, this implies that observations from the same cluster
are correlated. If this correlation is left unaddressed, estimates of
\emph{p} values, confidence intervals even model parameters are prone to
bias \citep{loca01}. {[}TODO: make a link to imputation methods, which
require adequate handling and propagation of variance; we are not
recommending the adoption of multilevel models for data analysis here,
but rather for imputation. VMTdJ: ``I would only make that link after
introducing missing data, and missing data is not mentioned yet, so I
would do that in the next section''{]} Statistical methods for clustered
data typically adopt hierarchical models that explicitly describe the
grouping of observations. These models are also known as `multilevel
models', `hierarchical models', `mixed effect models', `random effect
models', and in the context of time-to-event data as `frailty models'.
Table \ref{tab:clus} provides an overview of some key concepts in
multilevel modeling.

Like any other dataset, clustered datasets are prone to missing data.
Several strategies can be used to handle missing data, including
complete case analysis and imputation. We focus on the latter approach
and discuss statistical methods for replacing the missing data with one
or more plausible values. Afterwards, the completed data can be analyzed
as if they were completely observed. In contrast to single imputation
(where missing data are only replaced once), multiple imputation allows
to preserve uncertainty due to missingness and is therefore recommended
(c.f. Rubin 1976).

When clustered datasets are affected by missing values, we can
distinguish between two types of missing data: sporadic missingness and
systematic missingness \citep{resc13}. Sporadic missingness arises when
variables are missing for some but not all of the units in a cluster
\citep{buur18, jola18}. For example, it is possible that test results
are missing for several students in one or more classes. {[}TODO:
Provide an example for one of the case studies below.{]} When all
observations are missing within one or more clusters, data are
systematically missing. {[}TODO: Refer to Figure 1 and put
interpretation in the figure caption. VMTdJ: ``Mention in text that this
is sporadic missingness?''{]}

\hypertarget{introduction-1}{%
\section{Introduction}\label{introduction-1}}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2361}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7639}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
Sample unit & Units of the population from which measurements are taken
in a sample. \\
Hierarchical levels & Data are grouped into clusters at different
levels, observations belonging to the same cluster are expected to share
certain characteristics. \\
Fixed effect & Effects that are constant across all sample units,
e.g.~something that researchers control for and can repeat, such as the
administration of a drug. \\
Random effect & Effects that are a source of random variation in the
data, and whose levels are not fully sampled. e.g.~individuals are drawn
from a population of hospitals, here it is not possible to sample all
hospitals but drug effects could vary between hospitals. \\
Mixed effect & Includes fixed and random effects, e.g.~the fixed effect
would be the treatment effect of a drug and the random effect would be
the ID of the hospital where the patient is treated. Multilevel models
typically accommodate for variability by including a separate group mean
for each cluster e.g random intercept on hospitals. In addition to
random intercepts, multilevel models can also include random
coefficients and heterogeneous residual error variances across clusters
\citep[see e.g.][\citet{hox17} and \citet{jong21}]{gelm06}. \\
ICC & The variability due to clustering is often measured by means of
the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high
ICC would indicate that a lot of variability is due to the cluster
structure. \\
Stratified intercept & \\
\bottomrule()
\end{longtable}

\hypertarget{missingness-in-multilevel-data-1}{%
\subsection{Missingness in multilevel
data}\label{missingness-in-multilevel-data-1}}

@@

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/patterns-1}

@@

\caption[Missingness in multilevel data]{Missingness in multilevel data}\label{fig:patterns}
\end{figure}
\end{CodeChunk}

Imputation of missing data requires consideration of the mechanism
behind the missingness. Rubin proposed to distinguish between data that
are missing completely at random (MCAR), data that are missing at random
(MAR) and data that are missing not at random (MNAR; see Table
\ref{tab:miss}). For each of these three missingness generating
mechanisms, different imputation strategies are warranted \citet{yuce08}
and \citet{hox15}. We here consider the general case that data are MAR,
and expand on certain MNAR situations.

\begin{table}[tb]
\caption{Concepts in missing data methods}
\label{tab:miss}
\centering
\begin{tabular}{ll}
\hline
\textbf{Concept} & \textbf{Details}   \\
\hline
MCAR    & Missing Completely At Random, where the probability to be missing is equal \\
& across all data entries \\
MAR     & Missing At Random, where the probability to be missing depends on observed \\
& information \\
MNAR    & Missing Not At Random (MNAR), where the probability to be missing \\
& depends on unrecorded information, making the missingness non-ignorable \\
& [@rubi76; @meng94]. \\
& [TODO: add congeniality, but maybe in-text?] \\
\hline
\end{tabular}
\end{table}

The \proglang{R} package \pkg{mice} has become the de-facto standard for
imputation by chained equations, which iteratively solves the
missingness on a variable-by-variable basis. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}. However, commonly used imputation methods were not
designed for use in clustered data and usually generate observations
that are independent whereas multilevel data are dependent. For this
reason, we discuss how the \proglang{R} package \pkg{mice} can be used
to impute multilevel data.

{[}TODO: clarify why clustering is relevant during imputation, and why
this exposes the need for specialized imputation methods and more
attention during their implementation (``thou shall not simply run
\texttt{mice()} on any incomplete dataset'').{]} {[}TODO: Add that the
more the random effects are of interest, the more you need multilevel
imputation models.{]} {[}TODO: Add an overview of all possible predictor
matrix values in manuscript or \texttt{ggmice} legend.{]}

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice} in \proglang{R}. We provide practical guidelines and
code snippets for different missing data situations, including
non-ignorable mechanisms. For reasons of brevity, we focus on multilevel
imputation by chained equations with \pkg{mice} exclusively; other
imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are
outside the scope of this tutorial. {[}TODO: change this sentence
because we do not assume familiarity: ``Assumed knowledge includes basic
familiarity with multilevel imputation \citep[see e.g.][ and
\citet{grun18}]{audi18} and the \pkg{lme4} notation for multilevel
models (see Table \ref{tab:mod})''{]}.

\begin{itemize}
\tightlist
\item
  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
  with data that are MAR,][]{mice};
\item
  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
  studies with data that are MAR,][]{metamisc};
\item
  \texttt{hiv} from the \pkg{GJRM} package \citep[simulated data on HIV
  diagnoses, \(n = 6,416\) patients across \(N = 9\) regions with data
  that are MNAR,][]{GJRM}.
\end{itemize}

This tutorial is dedicated to readers who are unfamiliar with multiple
imputation. More experienced readers can skip the introduction (case
study 1) and directly head to practical applications of multilevel
imputation under MAR conditions (case study IMPACT) or under MNAR
conditions (case study HIV).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2222}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7778}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
MCAR & Missing Completely At Random, where the probability to be missing
is equal across all data entries \\
MAR & Missing At Random, where the probability to be missing depends on
observed information \\
MNAR & Missing Not At Random (MNAR), where the probability to be missing
depends on unrecorded information, making the missingness non-ignorable
\citep{rubi76, meng94}. \\
\bottomrule()
\end{longtable}

\hypertarget{aim-of-this-paper-1}{%
\subsection{Aim of this paper}\label{aim-of-this-paper-1}}

@@

{[}TODO: Add environment info, seed and version number(s) somewhere.{]}
Set up the R environment and load the necessary packages:

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)         # for imputation
R> library(miceadds)     # for imputation
R> library(ggmice)       # for visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel pooling
R> library(miceheckman)       # for imputation cf. heckman models
R> library(metamisc)     # for case study data
\end{CodeInput}
\end{CodeChunk}

@@

{[}TODO: explain case study{]}

In this section we'll go over the different steps involved with imputing
incomplete multilevel data with the R package mice. We consider the
simulated \texttt{popmis} dataset, which included pupils (\(n = 2000\))
clustered within schools (\(N = 100\)). The following variables are of
primary interest:

@@

The true effect is:

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/unnamed-chunk-3-1} \end{center}

@@

Plot the missing data pattern:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(popmis)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop_pat-1}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop-corr-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop-hist-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop-ignored-pred-1} \end{center}

@@

{[}TODO: Add that it doesn't work with systematic missingness (only with
sporadic). There's some pros and cons, and it may not even differ much
if the number of clusters is low.{]}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop_predictor-1} \end{center}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pop_multilevel-1} \end{center}

@@

{[}TODO: check if there is systematic missingness in this dataset, if
not make Marshall Computerized Tomography classification (ct)
systematically missing.{]}

We illustrate how to impute incomplete multilevel data by means of a
case study: \texttt{impact} from the \pkg{metamisc} package
\citep[empirical data on traumatic brain injuries, \(n = 11,022\) units
across \(N = 15\) clusters,][]{metamisc}. {[}TODO: add more info about
the complete data.{]} The \texttt{impact} data set contains traumatic
brain injury data on \(n = 11022\) patients clustered in \(N = 15\)
studies with the following 11 variables:

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification, {[}TODO:
  make this one var? also shows that you don't always need random
  effects everywhere?{]}
\item
  \texttt{hypox} Hypoxia (0=no, 1=yes),
\item
  \texttt{hypots} Hypotension (0=no, 1=yes),
\item
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
\item
  \texttt{edh} Epidural hematoma (0=no, 1=yes),
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/forest-1} \end{center}

@@

{[}TODO: show how much variance there is after different methods{]}

{[}TODO: add ICC before/after imputation and interpret: This tells us
that the multilevel structure of the data should probably be taken into
account. If we don't, we'll may end up with incorrect imputations,
biasing the effect of the clusters towards zero.{]}

{[}TODO: add descriptive statistics of the complete and incomplete
data.{]}

@@

To explore the missingness, it is wise to look at the missing data
pattern:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/pattern-1} \end{center}

@@

This shows\ldots{} {[}TODO: fill in that we need to impute \texttt{ct}
and \texttt{pupil}.{]}

To develop the best imputation model, we need to investigate the
relations between the observed values of the incomplete variables and
the observed values of other variables, and the relation between the
missingness indicators of the incomplete variables and the observed
values of the other variables. To see whether the missingness depends on
the observed values of other variables, we\ldots{} {[}TODO: fill in that
we can test this statistically or use visual inspection (e.g.~a
histogram faceted by the missingness indicator).{]}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/impact_corr-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect  group term  estimate std.error statistic   p.value conf.low conf.high
   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   3.00e-31   0.0571     0.130
 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.56e-19   1.92       2.76 
11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
\end{CodeOutput}
\end{CodeChunk}

As we can see\ldots{} {[}TODO: fill in: This means that a higher ct
(Marshall Computerized Tomography classification) is associated with a
lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI
\ldots{} to \ldots, when controlling for \ldots{]}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> meth <- make.method(dat) # methods vector
R> pred <- quickpred(dat)   # predictor matrix
R> plot_pred(pred)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/impact-1} \end{center}

\begin{CodeInput}
R> pred[pred == 1] <- 2
R> pred["mort", ] <- 2
R> pred[, "mort"] <- 2
R> pred[c("name", "type", "age", "motor_score", "mort"), ] <- 0
R> pred[, "name"] <- -2
R> diag(pred) <- 0
R> plot_pred(pred)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data2_files/figure-latex/impact-2} \end{center}

@@

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10505.026 10570.795 -5243.513 10487.026     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2908  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32339                 -0.42359                  0.03023  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67142                 -1.05776                 -1.51038  
                  pupil                       ct  
                0.49756                  0.42474  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10519.511 10585.280 -5250.755 10501.511     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2961  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.33581                 -0.40871                  0.03039  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66477                 -1.05451                 -1.51858  
                  pupil                       ct  
                0.45928                  0.44419  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10522.038 10587.807 -5252.019 10504.038     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2955  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.35187                 -0.40574                  0.03064  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67468                 -1.06551                 -1.52622  
                  pupil                       ct  
                0.47006                  0.44148  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings
